Data Migration under a Changing Schema

### 2023-07-06 2022Migration/articleMigrationFACS.tex
FACS article without conclusion.

\documentclass{elsarticle}
% \usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathrsfs}
\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\hyphenation{ExecEngine}
\newtheorem{lemma}{Lemma}
\begin{document}
\include{preambleMigrations}

% TODO Algemeen:
% - type van triple-set en atom-set definieren zodat viol_u ?->P(?x?)
%   (Bas: ik heb foute types weggehaald, maar dit nog niet gedefinieerd)
% - nieuwe populatie na de disjoint union zou de populatie met ENFORCE regels toegepast moeten zijn
%  (en in het bijzonder de ISA regels)

% TODO Stef:
% - stukje literatuuronderzoek
% https://ieeexplore.ieee.org/abstract/document/7445334?casa_token=ECzi6XeV2ncAAAAA:KhWzB8XBFOUJ0C6AD-XjX_ryuA9ARvTd3gm6RR-ZNiR8sZ1858FJpQ7zKQhkAZDlv8IjPdgD
% https://ieeexplore.ieee.org/abstract/document/8549944?casa_token=9qiGqNzh2Q0AAAAA:C-cYogExB35nGxQdxLcdBh4JoLNvM0OHedAMhCbB5V4kb4_6nzHUvc23xSJbeoBu67LSiz-Y
% https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.651.9298&rep=rep1&type=pdf
% https://www.scirp.org/html/4-7800724_106592.htm
% https://www.researchgate.net/profile/Ranjana-Badre/publication/318665687_GUI_for_Data_Migration_and_Query_Conversion/links/5b45bbea0f7e9b1c722386e5/GUI-for-Data-Migration-and-Query-Conversion.pdf
% https://journal3.uin-alauddin.ac.id/index.php/literatify/article/view/12567

% Bewijsverplichtingen:
% -> het migratie-systeem is typefout-vrij (opmerking: als we de type-checker buiten beschouwing willen laten, kunnen we dit niet beschrijven)
% -> het migratie-systeem is vrij van overtredingen op regels die het migratie-systeem moet bewaken
% -> het migratie-systeem bevat alle oude data, ihb nog steeds na het toepassen van de enforce regels
% -> er is een pad naar ingebruikname van het nieuwe systeem (vanaf de initiële toestand van het migratie-systeem)
% -> corollary: op het moment van ingebruikname van het nieuwe systeem, is het migratie-systeem vrij van overtredingen op regels die het nieuwe systeem moet bewaken
% -> optioneel: na een begrensd aantal 'voortgangs-stappen' kan het nieuwe systeem in gebruik genomen worden
% -> optioneel: er bestaat een migratie-systeem dat de oude functionaliteit behoudt (mogelijk uitbreid) totdat het nieuwe systeem in gebruik genomen is?

\title{Data Migration under a Changing Schema}
\author[ou,ordina]{Stef Joosten\fnref{fn1}}
\ead{stef.joosten@ou.nl}
\author[umn]{Sebastiaan Joosten\fnref{fn2}}
\address[ou]{Open Universiteit Nederland, Heerlen, the Netherlands}
\address[ordina]{Ordina NV, Nieuwegein, the Netherlands}
\address[umn]{University of Minnesota, Minneapolis, USA}
\fntext[fn1]{ORCID 0000-0001-8308-0189}
\fntext[fn2]{ORCID 0000-0002-6590-6220}

\begin{abstract}
   Software generators can help to increase the frequency of releases and their reliability.
   They save on time spent on development and time spent on fixing human-induced mistakes by compiling a specification into a working information system.
   However, many generators do not support data migrations.
   A data migration is necessary when an incremental deployment changes the schema of the system.
   As a consequence, developers tend to avoid migrations or migrate data ``by hand''.

   To address this problem, this paper proposes a theory for data migrations aimed at automating the migration process.
   The problem at large is how to preserve the semantics of that data under a changing schema.
   This paper proposes a theory for deploying an incremental change.
   The theory is applicable in general, but will be implemented in a software generator called Ampersand.

   This paper aims to preserve the semantics of data by satisfying concrete business requirements.
   The migration process is based on the assumptions that
   software is deployed incrementally,
   the existing data set may be polluted,
   human interaction may be required,
   the meaning of data must be preserved,
   the business continues during the migration without interruption (zero down-time),
   and there is a compiler to generate an information system from a given schema.
   The correctness of the migration is the focus of this paper, while efficiency is outside its scope.
\end{abstract}

\begin{keyword}
generative software\sep incremental software deployment\sep data migration\sep relation algebra\sep Ampersand\sep schema change
\end{keyword}
\maketitle

\section{Introduction}
\label{sct:Introduction}
   The purpose of this work is to automate the data migrations that come with incrementally developing and deploying information systems.
   This paper proposes a theory for such data migrations in the context of generated information systems%
\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }.

   This paper is founded in the belief that software generators help developers to increase the frequency of deploying successive increments.
   They also prevent some human-induced errors, both in the process and in the resulting software.
   We believe that software generation can thus lead to a higher pace of deployments, a more predictable deployment process, and more reliable information systems.

   A problem is that software generators typically do not support the data migration process, especially when the schema changes.
   The ``manual'' effort required to migrate data slows down the development process and introduces mistakes%
\footnote{These effects can be measured in terms of DevOps metrics such as
   deployment frequency,
   reliability of deployments, and
   change failure rate~\cite{DevOps2021}.}.

   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{JoostenRAMiCS2017, Joosten-JLAMP2018}.
   It aims at automating data migrations, so incremental changes can be done more reliably and faster.
   This contributes to a more agile software development process.
   We believe that users will experience a more organic evolution of the system because increments are smaller, occur more frequently and more reliably.

@@

   Data migration for other purposes has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers may deliberately avoid schema differences and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
   For example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define an increment as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   Obviously, such data migrations are not suitable for incremental software deployment,
   if the schema changes and the old functionality has to be replaced.
   Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
   deployed in increments.

@@

\section{Analysis}
\label{sct:Analysis}
\subsection{Information Systems}
   The purpose of an information system is to make data meaningful to its users.
   The analysis of the problem starts with a formal definition of information systems,
   in which the schema contains all information that makes data meaningful.
   We then define a migration procedure from one information system to another,
   which preserves the meaning of the data as much as possible.

   Every user has her\footnote{Throughout this paper, we ask the reader to read ``she'' or ``her'' to include the masculin form as well.} own tasks and responsibilities
   and may work from different locations and on different moments.
   This collective use by multiple users serves a purpose which we will loosely call ``the business''.
   It is this business that depends on the semantics of the data to draw the right conclusions and carry out their tasks.
   This paper uses semantic constraints to represent this meaning.

   Information systems are typically used by actors (both users and computers) who are distributed and work with data all the time.
   As a consequence, the data in a system changes continually.
   In practice, actors ``talk to'' the system through an ingress mechanism, which connects each user to the right service(s).
   The ingress function is provided by the deployment platform, so it is beyond the scope of this paper.
   Also, every service runs independent of other services,
   meaning that each service can be stopped, (re)started and substituted without disrupting the other services.
   For practical purposes, this presumes a platform of managed services, such as Kubernetes,
   even though the theory in this paper is not specific to any platform.

   Every system contains a data set, which represents the state of the system.
   Every service produces and consumes events that may change the state of the system.
   This state is represented in a persistent store, aka the database%
\footnote{Whether the database is local, remote, or globally distributed is immaterial for the theory in this paper.}.
   Events that the system detects may cause the state to change.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
   \end{center}
\caption{Anatomy of an information system}
\label{fig:pre-migration}
\end{figure}
   To keep our theory technology independent, data sets are assumed to contains triples.
   This makes our theory valid for any kind of database that triples can represent,
   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
   The system's semantics is represented as constraints,
   which most database management systems refer to as integrity rules.
   The purpose of these rules is to keep all constraints satisfied.

   Like most other software generators, Ampersand is a compiler that generates information systems from a script.
   Ampersand uses a syntactically sugared form of heterogeneous relation algebra,
   which works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
\footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
   Each rule in Ampersand is a constraint (aka invariant), which the system keeps satisfied as long as that rule lives%
\footnote{The dotted lines in the diagrams show which set of rules is enforced.}.
   So, developers spend no effort to extract constraints from specifications or (even more laborious) from code
   because all constraints are explicitly available as rules in the code.

@@

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.

   In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
   while transferring data in a controlled fashion.
   Typically, a migration strategy (e.g. case by case, customer by customer, or in batches) is in place
   to ensure the preservation of valuable data and the quality of the migration.
   To automate this process completely is unrealistic in many cases%
\footnote{Note that smaller increments improve the chance that a fully automic migration is possible,
although there will always be cases where human intervention is required.}.
   Manual interventions can be necessary to resolve data pollution, new business rules, or to fix known issues in the old system.
   Consequently, a migration engineer must have the freedom to change the migration script to suit the particulars of the migration.
   Such complications may arise mostly as a result of data pollution.
   We distinguish data pollution in the following categories:
\begin{itemize}
   \item data pollution that violates a constraint in the schema.
   \item data pollution that violates a constraint that is not in the schema.
   \item data pollution that cannot be captured in constraints on the data set,
   such as a street address that has become obsolete because someone has moved without notification.
   \item data pollution that is a consequence of an erroneous constraint in the schema.
\end{itemize}
   The first category is about violations of constraints from the schema.
   Such violations do not occur in our theory because the existing system was generated by Ampersand,
   so the data set satisfies all constraints in the schema.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   The second category, data pollution that is not captured by any of the constraints mentioned in the schema,
   can be dealt with by adding constraints in the desired system.
   In that case, a migration engineer must ensure that the desired system can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary in specific cases.
   The third category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   So, such constraints may help to move this type of pollution to one of the other categories.
   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the desired system.

   This illustrates that automating data migrations may require user intervention,
   either by a migration engineer or by end-users.
   In our approach, such user interventions are given some time by
   defining two distinct moments:
\begin{enumerate}
   \item the moment of transition (the MoT), i.e. the moment that changed functionality is made available to users;
   \item the moment of cleanup (the MoC), i.e. the moment after which the old software and data can be removed.%
\footnote{Backup mechanisms are outside the scope of this paper.}.
\end{enumerate}

   During the time between the MoT and MoC,
   the existing system and the desired system are kept alive, side by side, as shown in figure~\ref{fig:migration phase}.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{figures/datamigration-Migration phase.png}
   \end{center}
\caption{Migration phase}
\label{fig:migration phase}
\end{figure}

   The migration requires a third schema, which specifies the migration itself.
   The migration schema comprises the schemas of both the existing system and the desired system.
   The rules of the migration schema will cause the right data to be copied correctly into the desired system.
   This migration schema preserves data from the existing system, it may introduce new data automatically (generated from the existing data set),
   and it may require users to introduce new data (which will take time). 
   So the idea of a migration schema is to compute the disjoint union of the existing and the desired schemas
   and to present it to the migration engineer in the form of source code,
   so the migration engineer can change everything she wants to suit the particulars of the data migration.

   During the migration phase, transactions in the existing system are allowed because the migration engine will transport them to the desired system.
   This allows the business to finish transactions in the existing system while the desired system is already up and running.
   Similarly, the existing system must execute transactions from the desired system, just in case the business calls off the migration.
   Rules in the migration system that require user interaction are given time by requiring that the migration phase ends
   only after all migration rules are at rest.
   At the MoC, the migration engineer takes down everything but the desired system,
   leaving the business with a successful migration.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{figures/datamigration-Post-migration.png}
   \end{center}
\caption{The system after the data migration}
\label{fig:post-migration}
\end{figure}

@@

\section{Terminology}
\label{sct:Terminology}
   An {\em information system} is a combination of data set, schema, and functionality.
   For the purpose of this paper, the functionality is ignored because it does not impact the migration.
   Section~\ref{sct:Data sets} desribes data sets. Schemas are treated in section~\ref{sct:Schemas}.
   Then section~\ref{sct:Information Systems} defines information systems.

@@

   Atoms serve as data elements.
   Atoms are values without internal structure of interest, meant to represent atomic data elements (e.g. dates, strings, numbers, etc.) in a database.
   From a business perspective, atoms represent concrete items of the world,
   such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
   By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
   All atoms are taken from an infinite set called $\Atoms$.

   Concepts are names that group atoms of the same type.
   All concepts are taken from an infinite set $\Concepts$.
   $\Concepts$ and $\Atoms$ are disjoint.
   For example, a developer might choose to classify \atom{Peter} and \atom{Melissa} as \concept{Person},
   and \atom{074238991} as a \concept{TelephoneNumber}.
   In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
   In the sequel, variables $A$, $B$, $C$, $D$ will represent concepts.

   The relation $\inst:\Pair{\Atoms}{\Concepts}$ relates atoms to concepts.
   The term $a\inst C$ means that atom $a$ is an \emph{instance} of concept $C$.
   This relation is used in the type system, in which $\inst$ assigns one or more concepts to every atom in the data set.
   Since $\inst$ is a relation and every relation is a set of pairs,
   set operators $\cup$, $\cap$, and $-$ can be used on $\inst$.

%    Specialization, also known as {\em generalization} or {\em subtyping}, is a relation between concepts.
%    The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
% \begin{equation}
%    \label{eqn:specialization}
%    A\isa B\ \Leftrightarrow\ \forall a: a\inst A\rightarrow a\inst B
% \end{equation}
%    Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
%    In her script, a user can declare a specialization as a pair of concepts in the relation $\isa$.
%    A compiler can construct $\isa$ as the transitive, reflexive closure of all user-defined specializations.
%    It must make sure that $\isa$ is antisymmetric, so that $\isa$ is a partial order of concepts.
%    Specialization causes atoms to have multiple concepts of which they can be an instance.
%    As a result, if an atom is an instance of concept $A$ and $A\isa B$,
%    this atom has all properties that atoms of type $B$ have.

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%

\usepackage{hyperref}
%\usepackage{multicol}
%\usepackage{footmisc}
%\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathrsfs}
\usepackage{mathtools}
%\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\hyphenation{ExecEngine}
%\newtheorem{lemma}{Lemma}

% Ampersand -----------------------------------------------------------

%\def\id#1{\text{\it #1\/}}
\newcommand{\xrightarrowdbl}[2][]{%
  \xrightarrow[#1]{#2}\mathrel{\mkern-14mu}\rightarrow
}
\newcommand{\id}[1]{\text{\it #1\/}}
\newcommand{\code}[1]{\text{\tt\small #1}}
\newcommand{\stmtText}[1]{``{\small\tt #1}''}
\newcommand{\dom}[1]{\id{dom}(#1)}
\newcommand{\cod}[1]{\id{cod}(#1)}
%\renewcommand{\int}[2]{\id{inter}(#1,#2)}
\newcommand{\relsIn}[1]{\id{relsIn}(#1)}    % maps a Term to a set of Relations
\newcommand{\pop}[2]{\id{pop}_{#1}(#2)}
\newcommand{\maintain}{\mathbin{\id{maint}}}
\newcommand{\enf}{\mathbin{{\tt enforce}}}
\newcommand{\enforce}[2]{{\tt enforce}_{#1}(#2)}
\newcommand{\instance}{\mathbin{\id{inst}}}
\newcommand{\relname}[1]{\id{relname}(#1)}
\newcommand{\evt}[2]{\id{event}_{#1,#2}}
\newcommand{\src}[1]{\id{src}(#1)}
\newcommand{\tgt}[1]{\id{tgt}(#1)}
\newcommand{\sat}[2]{\id{sat}_{#1}(#2)}
\newcommand{\viol}[2]{\violC{#1}(#2)}
\newcommand{\violC}[1]{\id{viol}_{#1}}
\newcommand{\sign}[1]{\id{sign}(#1)}
\newcommand{\powerset}[1]{\cal{P}\{#1\}}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\full}{V}
\newcommand{\declare}[3]{\id{#1}_{\pair{#2}{#3}}}
\newcommand{\subst}[3]{#3_{[#1\rightarrow #2]}}
\newcommand{\fullt}[2]{V_{\pair{#1}{#2}}}
\newcommand{\iden}{I}
\newcommand{\ident}[1]{I_{\id{#1}}}
\newcommand{\expr}[3]{(#1)_{#2\times #3}}
\newcommand{\pair}[2]{\langle{#1},{#2}\rangle}
\newcommand{\maprel}[2]{{\tt maprel}_{#1}({#2})}
\newcommand{\Pair}[2]{#1\times#2}
\newcommand{\pairs}[1]{\id{pairs}(#1)}
\newcommand{\triple}[3]{\langle{#1},{#2},{#3}\rangle}
\newcommand{\quadruple}[4]{\langle{#1},{#2},{#3},{#4}\rangle}
\newcommand{\atom}[1]{{\tt\small #1}}
\newcommand{\atoms}{\mathcal{A}}
\newcommand{\Atoms}{\mathbb{A}}
\newcommand{\events}{\mathcal{E}}
\newcommand{\Events}{\mathbb{E}}
\newcommand{\concept}[1]{{\tt\small #1}}
\newcommand{\concepts}{\mathcal{C}}
\newcommand{\Concepts}{\mathbb{C}}
\newcommand{\decls}{\mathcal{D}}  %% names of relations
\newcommand{\rels}{\mathcal{R}}   %% all relations
\newcommand{\Rels}{\mathbb{R}}   %% all relations
\newcommand{\relations}{\mathcal{M}} % representing terms. M is a subset of R.
\newcommand{\triples}{\mathcal{T}}
\newcommand{\Triples}{\mathbb{T}}
\newcommand{\Triple}[3]{#1\times#2\times#3}
\newcommand{\vertices}{N}
\newcommand{\rules}{\mathcal{U}}
\newcommand{\Rules}{\mathbb{U}}
\newcommand{\specrules}{\mathcal{S}}
\newcommand{\roles}{\mathcal{O}}
\newcommand{\dataset}{\mathscr{D}}
\newcommand{\Dataset}{\mathbb{D}}
\newcommand{\schema}{\mathscr{Z}}
\newcommand{\functionality}{\mathscr{F}}
\newcommand{\select}[2]{\id{select}_{#1}\{{#2}\}}
\newcommand{\migrsys}{\mathscr{M}}
\newcommand{\infsys}{\mathscr{S}}
\newcommand{\tf}[1]{\mathscr{T}(#1)}
\newcommand{\ptf}[1]{\mathscr{T}'(#1)}
\newcommand{\ti}[1]{\mathscr{I}(#1)}
\newcommand{\tic}[1]{I_{\cal C}(#1)}
\newcommand{\relAdd}{\dagger}
\newcommand{\flip}[1]{{#1}^\smallsmile} %formerly:  {#1}^\backsim
\newcommand{\kleeneplus}[1]{{#1}^+}
\newcommand{\kleenestar}[1]{{#1}^*}
\newcommand{\cmpl}[1]{\overline{#1}}
\newcommand{\rel}{\times}
\newcommand{\compose}{;}
\newcommand{\subs}{\subseteq}%{\models}
\newcommand{\fun}{\rightarrow}
\newcommand{\isa}{\preceq}
%\newcommand{\isaClos}{\sqsubseteq}
\newcommand{\typetest}{?}
\newcommand{\meet}{\sqcap}
\newcommand{\join}{\sqcup}
\newcommand{\Meet}{\bigsqcap}
\newcommand{\Moin}{\bigsqcup} % because LaTeX has already defined command \Join.
\newcommand{\order}{\ominus}
\newcommand{\anything}{\top}
\newcommand{\nothing}{\bot}
\newcommand{\rewriteto}{\rightarrow}
\newcommand{\calc}{\implies}
\newcommand{\alland}{\bigwedge}
\newcommand{\mph}[3]{#1_{#2\times #3}}
\newcommand{\mphu}[1]{#1_{\univ\times\univ}}

%-----------------------------------------
\newcommand{\kse}{\hspace*{1.7em}}
\newcommand{\ksf}{\hspace*{1em}}
\newcommand{\ksg}{\hspace*{1em}}
\newenvironment{derivation}{\begin{tabbing}\kse \= \ksf \= \ksg \= \kill}{\end{tabbing}}
%\newtheorem{definition}{Definition}
\newcommand{\term}[1]{\>\>\(#1\)\\[1ex]}
\newcommand{\rela}[2]{\>\(#1\)\>\>\{ \ #2 \ \}\\[1ex]}
\newcommand{\weg}[1]{}

\def\define#1{\label{dfn:#1}{\em #1}\index{#1}}
\def\definem#1{\label{dfnm:#1}{\em #1}\index{#1}\marge{#1}}
\newcommand{\marg}[1]{\index{#1}\marge{#1}}

\usepackage{color}
\renewcommand\UrlFont{\color{blue}\rmfamily}
%

\begin{document}
%

\title{Data Migration under a Changing Schema}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sebastiaan Joosten\inst{1}\orcidID{0000-0002-6590-6220}\\ \and
Stef Joosten\inst{2,3}\orcidID{0000-0001-8308-0189}}
%
\authorrunning{S. Joosten}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of Minnesota, Minneapolis, USA \and Open Universiteit Nederland, Heerlen, the Netherlands\\ 
\email{stef.joosten@ou.nl} \and
Ordina NV, Nieuwegein, the Netherlands}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Software generators can help to increase the frequency of releases and their reliability. They save on time spent on development and fixing human-induced mistakes by compiling a specification into a working information system. However, many generators do not support data migrations. Data migration is necessary when an incremental deployment changes the system's schema. Consequently, developers tend to avoid migrations or migrate data ``by hand''.

To address this problem, this paper proposes a theory for data migrations aimed at automating the migration process. The problem at large is how to preserve the semantics of that data under a changing schema. This paper proposes a theory for deploying incremental change. The theory is applicable in general but will be implemented in a software generator called Ampersand.

This paper aims to preserve the semantics of data by satisfying concrete business requirements. The migration process is based on the assumption that software is deployed incrementally, the existing data set may be polluted, human interaction may be required, the meaning of data must be preserved, the business continues during the migration without interruption (zero downtime), and there is a compiler to generate an information system from a given schema. The correctness of the migration is the focus of this paper, while efficiency is outside its scope.

\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sct:Introduction}
   The purpose of this work is to automate the data migrations that come with incrementally developing and deploying information systems.
   This paper proposes a theory for such data migrations in the context of generated information systems%
\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }.

   A problem is that software generators typically do not support the data migration process, especially when the schema changes.
   The ``manual'' effort required to migrate data slows down the development process and introduces mistakes%
   Such effects can be measured in terms of DevOps metrics such as deployment frequency, reliability of deployments, and change failure rate~\cite{DevOps2021}.

   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}.
   It aims at automating data migrations, so incremental changes can be done more reliably and faster.
   This contributes to a more agile software development process.
   We believe that users will experience a more organic evolution of the system because increments are smaller, occur more frequently and more reliably.

@@

   Data migration for other purposes has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers may deliberately avoid schema differences and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
   For example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define an increment as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   Such data migrations do not allow for schema changes and do not fit our need for incremental software deployment.
   Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
   deployed in increments.

@@

\section{Analysis}
\label{sct:Analysis}
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   This section defines information systems, after which we define a migration procedure from one information system to another.

   Every user has her own tasks and responsibilities and may work from different locations and on different moments.
   This collective use by multiple users serves a purpose which we will loosely call ``the business''.
   It is this business that depends on the semantics of the data to draw the right conclusions and carry out their tasks.
   This paper uses semantic constraints to represent this meaning.

   Information systems are typically used by actors (both users and computers) who are distributed and work with data.
   As a consequence, the data in a system changes continually.
   The state of the system is represented by a data set, typically represented in a persistent store or database.
   Events that the system detects may cause the state to change.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
   \end{center}
\caption{Anatomy of an information system}
\label{fig:pre-migration}
\end{figure}
   To keep our theory technology independent, data sets are assumed to contain triples.
   This makes our theory valid for any kind of database that triples can represent,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
   An important aspect of the system's semantics are the constraints, sometimes known as invariants or integrity rules.
   %The purpose of these rules is to keep all constraints satisfied.

   Like most software generators, Ampersand is a compiler that generates information systems from a script.
   Ampersand uses a form of heterogeneous relation algebra,
   which works with relations in a way similar to Alloy~\cite{Alloy2006}.
   A system is then described by a set of rules (aka constraint / invariant), which the system keeps satisfied.
   So all constraints are explicitly available as rules in the Ampersand code.

@@

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   %Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
   %
   In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
   while transferring data in a controlled fashion.
   Typically, a migration strategy (e.g. case by case, customer by customer, or in batches) is in place
   to ensure the preservation of valuable data and the quality of the migration.
   Automating this process completely is unrealistic in many cases:
   While smaller increments improve the chance that a fully automatic migration is possible, there will always be cases where human intervention is required.
   Manual interventions can be necessary to resolve data pollution, new business rules, or to fix known issues in the old system.
   Consequently, a migration engineer must have the freedom to tailor any automatic approach to migration to suit the particulars.

   Complications that require manual intervention arise mostly as a result of data pollution.
   We distinguish data pollution in the following categories:
\begin{itemize}
   \item data pollution that violates a constraint in the schema.
   \item data pollution that violates a constraint that is not in the schema.
   \item data pollution that cannot be captured in constraints on the data set,
   such as a street address that has become obsolete because someone has moved without notification.
   \item data pollution that is a consequence of an erroneous constraint in the schema.
\end{itemize}
   The first category is about violations of constraints from the schema.
   Such violations do not occur in our theory because we assume the existing system was generated by Ampersand,
   so the data set satisfies all constraints in the schema.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   The second category, data pollution that is not captured by any of the constraints mentioned in the schema,
   can be dealt with by adding constraints in the desired system.
   In that case, a migration engineer must ensure that the desired system can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary in specific cases.
   The third category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   So, such constraints may help to move this type of pollution to one of the other categories.
   The last category of data pollution, erroneous constraints, must be fixed by removing them or replacing them by the correct constraints in the desired system.

   This illustrates that automating data migrations may require user intervention,
   either by a migration engineer or by end-users.
   In our approach, such user interventions are given some time by
   defining two distinct moments:
\begin{enumerate}
   \item the moment of transition (the MoT), i.e. the moment that changed functionality is made available to users;
   \item the moment of cleanup (the MoC), i.e. the moment after which the old software and data can be removed.
   This leaves the business with the intended system: a successful migration.
\end{enumerate}

   During the time between the MoT and MoC,
   the existing system and the desired system are kept alive, side by side, as shown in Figure~\ref{fig:migration phase}.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{figures/datamigration-Migration-phase.png}
   \end{center}
\caption{Migration phase}
\label{fig:migration phase}
\end{figure}

   The migration requires a third schema, which specifies the migration itself.
   The migration schema comprises the schemas of both the existing system and the desired system.
   The rules of the migration schema will cause the right data to be copied correctly into the desired system.
   This migration schema preserves data from the existing system, it may introduce new data automatically (generated from the existing data set),
   and it may require users to introduce new data (after the MoT and before the MoC).
   So the migration schema is a (disjoint) union of the existing and the desired schemas,
   which is presented to the migration engineer in the form of source code,
   so the migration engineer can change everything she wants to suit the particulars of the data migration.

%During the migration phase, transactions in the existing system are allowed because the migration engine will transport them to the desired system.
%This allows the business to finish transactions in the existing system while the desired system is up and running.
%Similarly, the existing system must execute transactions from the desired system, just in case the business calls off the migration.
%Rules in the migration system that require user interaction are given time by requiring that the migration phase ends
%only after all migration rules are at rest.

\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{figures/datamigration-Post-migration.png}
   \end{center}
\caption{The system after the data migration}
\label{fig:post-migration}
\end{figure}

@@

\section{Terminology}
\label{sct:Terminology}
   An {\em information system} is a combination of data set, schema, and functionality.
   For the purpose of this paper, functionality captured in user interfaces are ignored because it does not impact the migration.
   Section~\ref{sct:Data sets} desribes data sets. Schemas are treated in section~\ref{sct:Schemas}.
   Then section~\ref{sct:Information Systems} defines information systems.

@@

   Atoms serve as data elements.
   %Atoms are values without internal structure of interest, meant to represent atomic data elements (e.g. dates, strings, numbers, etc.) in a database.
   %From a business perspective, atoms represent concrete items of the world,
   %such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
   %By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
   All atoms are taken from an infinite set called $\Atoms$.
   %
   Concepts are names that group atoms of the same type.
   All concepts are taken from an infinite set $\Concepts$.
   %$\Concepts$ and $\Atoms$ are disjoint.
   For example, a developer might choose to classify \atom{Peter} and \atom{Melissa} as \concept{Person},
   and \atom{074238991} as a \concept{TelephoneNumber}.
   In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
   Moreover, \atom{Peter}, \atom{Melissa} and \atom{074238991} are atoms.
   In the sequel, variables $A$, $B$, $C$, $D$ will represent concepts, and variables $a$, $b$, and $c$ represent \emph{atoms}.
   %
   The relation $\instance:\Pair{\Atoms}{\Concepts}$ relates atoms to concepts.
   The term $a\instance C$ means that atom $a$ is an \emph{instance} of concept $C$.
   %This relation is used in the type system, in which $\instance$ assigns one or more concepts to every atom in the data set.
   %Since $\instance$ is a relation and every relation is a set of pairs,
   %set operators $\cup$, $\cap$, and $-$ can be used on $\instance$.

%    Specialization, also known as {\em generalization} or {\em subtyping}, is a relation between concepts.
%    The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
% \begin{equation}
%    \label{eqn:specialization}
%    A\isa B\ \Leftrightarrow\ \forall a: a\instance A\rightarrow a\instance B
% \end{equation}
%    Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
%    In her script, a user can declare a specialization as a pair of concepts in the relation $\isa$.
%    A compiler can construct $\isa$ as the transitive, reflexive closure of all user-defined specializations.
%    It must make sure that $\isa$ is antisymmetric, so that $\isa$ is a partial order of concepts.
%    Specialization causes atoms to have multiple concepts of which they can be an instance.
%    As a result, if an atom is an instance of concept $A$ and $A\isa B$,
%    this atom has all properties that atoms of type $B$ have.
